---
title: "Avoid vignette code testing on CRAN and Travis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vignettes on CRAN and Travis-ci

By default when uploading a package to CRAN or testing a package on Travis, the code on the 
vignette will be run in order to make sure that it is reproducible. That is understandable and 
actually really helpful, as it minimises bugs. 

## The problem

What happens though when your code either takes too long to run (imagine running a complicated
model which takes 20 hours) or requires using a password (imagine creating an API where you would 
need to use a username and password) ? In the first case you wouldn't want to hammer CRAN's 
or Travis's servers whereas in the second case you wouldn't want to show your credentials to 
everyone reading the vignette. However, when building the vignette locally you would definitely want to 
test that your code works. This leads us to the question:

How can we ignore vignette code testing on CRAN or Travis, but test it locally?

## The solution - CRAN

When I was preparing the RDota2 package (an API client for Valve's Dota2) I needed to use a key to
access the API. In order to do this I created an environment variable with the key value locally. 
Every time I wanted to test the package, I would load this environment variable  
(with `Sys.getenv('RDota_KEY')`) and use it to log in to the API. Obviously, the environment 
variable was saved on my local machine. When testing my package with `devtools::check` the 
environment variable would be loaded and all of the code would run fine. However, this would 
cause an error on CRAN and Travis, as they wouldn't be able to find the environment variable. 
(For information on how to set up an environment variable please check the Appendix at the end of 
this article).

Therefore, I needed to find a way to test my vignette code locally (making sure that the functions
work correctly with my key) but not on CRAN and Travis. For CRAN I used the following 
(modified from `httr`'s tutorial):

If we setup an environment variable on our local machines named `CRAN` and set that to `FALSE`,
then we can use that in a setup chunk in the vignette:

```{r, eval = FALSE}
CRAN <- !identical(tolower(Sys.getenv("CRAN")), "true")
knitr::opts_chunk$set(purl = CRAN)
```

and then you can use `eval = CRAN` as a chunk option (in next chunks).

When building the package locally `!identical(tolower(Sys.getenv("CRAN")), "true")` will return
`TRUE`. This will result in `eval = TRUE` in all chunks and the code will be checked. On the other 
hand, `!identical(tolower(Sys.getenv("CRAN")), "true")` will return `FALSE` on CRAN, 
which will result in `eval = FALSE` in individual chunks and the code will not run.

## The solution - Travis

Although the Cran solution above would work for Travis too (since there is no `CRAN` environment
variable), I also provide a secondary solution in case anyone needs it.

On Travis, by default, there is an environment variable called `Travis`. `Travis` is set to `true`. 
Using the same technique as CRAN we specify in our setup chunk:

```{r, eval = FALSE}
TRAVIS <- !identical(tolower(Sys.getenv("TRAVIS")), "true")
knitr::opts_chunk$set(purl = TRAVIS)
```

and then you can use `eval = TRAVIS` as a chunk option (in next chunks).

When building the package locally `!identical(tolower(Sys.getenv("TRAVIS")), "true")` will return
`TRUE`. This will result in `eval = TRUE` in all chunks. On the other hand, 
`!identical(tolower(Sys.getenv("TRAVIS")), "true")` will return `FALSE` on TRAVIS, 
which will result in `eval = FALSE` in individual chunks and the code will not run.

## The solution - Travis and CRAN

We can actually just use `CRAN` in the same way defined in the CRAN solution as this checks the
local `CRAN` environment variable we defined (this won't exist on CRAN or TRAVIS).

## Apendix - How to setup an environment variable

There are 5 easy steps in order to create an environment variable. I demonstrate the steps from 
the RDota2 package, but the methodology is exactly the same:

1. Identify your home directory. If you don't know which one it is just run `normalizePath("~/")` in the R console. 
2. In your home directory create a file called .Renviron (it shouldn't have an extension, like for example .txt). If questioned, YES you do want to use a file name that begins with a dot. Note that by default dotfiles are usually hidden. But within RStudio, the file browser will make .Renviron visible and therefore easy to edit in the future.
3. In the .Renviron file type a line like `RDota_KEY=xxxxxxxx`, where RDota_KEY will be the name of the R environment variable and xxxxxxxx will be your individual API Key. Make sure the last line in the file is empty (if it isn't R will silently fail to load the file). If you're using an editor that shows line numbers, there should be two lines, where the second one is empty.
4. Restart your R session if you were using one, since .Renviron is parsed at the start of an R session.
5. Access the key on your R session using `Sys.getenv`.