---
title: ""
output: html_document
---

## Beautiful HTML tables in R with tableHTML

A couple of years ago while I was developing a [football app](https://lyzander.shinyapps.io/FootballeR/) using shiny, I found 
it somewhat difficult to build a pretty HTML table exactly the way I wanted. 
For example, drawing vertical lines every three columns seemed like a difficult task back then even using
great packages like `xtable`. This provided the motivation to team up
with [Clemens](https://uk.linkedin.com/in/clemenszauchner/en) and develop `tableHTML`.

`tableHTML` is used to make CSS-ible HTML tables in an intuitive way. Those tables can
be used in shiny applications, Rmarkdown documents (like this one!) or any 
other application that accepts HTML. 

Let's see a few of its main uses.

## A simple HTML table

The main `tableHTML` function creates a blank HTML table on which CSS can be applied afterwards with the `add_css` functions. The package uses the pipe operator `%>%` coming from `magrittr` to join functions together.

```{r simple example, message=FALSE}
library(tableHTML)
mtcars[1:10, ] %>%
  tableHTML(widths = c(140, rep(50, 11)))
```

## A scientific HTML table

`tableHTML` provides two themes (rshiny-blue and scientific) that can be used directly with the `tableHTML` function.

```{r scientific}
library(tableHTML)
mtcars[1:10, ] %>%
  tableHTML(widths = c(140, rep(50, 11)),
            theme = 'scientific')
```

## Features of tableHTML

`tableHTML` allows the user to add to the table:

* New headers
* Second headers
* Row names
* Row groups
* A title
* A footer

All these can be specified in `tableHTML`'s arguments.

```{r caption headers}
mtcars[1:15, ] %>%
  tableHTML(widths = c(120, 140, rep(45, 11)),
            second_headers = list(c(2, 2, 4, 5), c('', 'team1', 'team2', 'team3')),
            row_groups = list(c(7, 8), c('Group1', 'Group2')),
            caption = 'This is a caption',
            footer = 'This is a footer')
```


## Beautify with CSS

Currently the package offers the following `add_css` functions which add CSS to their respective part of the HTML table:

* add_css_caption    
* add_css_column
* add_css_conditional_column
* add_css_footer             
* add_css_header            
* add_css_row              
* add_css_second_header  
* add_css_table              
* add_css_tbody             
* add_css_thead

This is an example of how some of the `add_css` functions can be used.

```{r add_css examples}
mtcars[1:15, ] %>%
  tableHTML(widths = c(140, rep(45, 11)),
            second_headers = list(c(3, 4, 5), c('team1', 'team2', 'team3')),
            caption = 'Table of Cars',
            footer = 'Figure 1. Stats for famous cars') %>% 
  add_css_second_header(css = list(c('height', 'background-color', 'font-size'), 
                                   c('40px', ' #e6e6e6', '30px')),
                        second_headers = 1:3) %>%
  add_css_header(css = list(c('height', 'background-color'), c('30px', ' #e6e6e6')),
                 headers = 1:12) %>%
  add_css_row(css = list('background-color', '#f2f2f2'),
              rows = even(1:17)) %>%
  add_css_row(css = list('background-color', '#e6f0ff'),
              rows = odd(1:17)) %>%
  add_css_column(css = list('text-align', 'center'), 
                 columns = names(mtcars)) %>%
  add_css_caption(css = list(c('text-align', 'font-size', 'color'), c('center', '20px', 'black'))) %>%
  add_css_footer(css = list(c('text-align', 'color'), c('left', 'black')))

```

## Conditional Formatting

Another special feature of the package is conditional formatting. You can see an example below using `add_css_conditional_column`. 

```{r}
mtcars[1:15, ] %>%
  tableHTML(widths = c(140, rep(45, 11))) %>%
  add_css_conditional_column(conditional = "==",
                             value = 14.3, 
                             css = list('background-color', "steelblue"), 
                             columns = 1) %>%
  add_css_conditional_column(conditional = "!=", 
                             value = 8,
                             css = list('background-color', "mediumvioletred"), 
                             columns = 2) %>%
  add_css_conditional_column(conditional = ">",
                             value = 440,
                             css = list('background-color', "orange"),
                             columns = 3) %>%
  add_css_conditional_column(conditional = ">=", 
                             value = 264,
                             css = list('background-color', "green"),
                             columns = 4) %>%
  add_css_conditional_column(conditional = "<",
                             value = 3, 
                             css = list('background-color', "yellow"),
                             columns = 5) %>%
  add_css_conditional_column(conditional = "<=", 
                             value = 2.20, 
                             css = list('background-color', "lightgray"), 
                             columns = 6)
```

## Colour ranking

Various colour rank themes (including custom) are also available.

```{r}
mtcars[1:15, ] %>%
  tableHTML(widths = c(140, rep(45, 11))) %>%
  add_css_conditional_column(colour_rank_theme = "RAG", columns = 1) %>%
  add_css_conditional_column(colour_rank_theme = "Spectral", columns = 2) %>%
  add_css_conditional_column(colour_rank_theme = "Rainbow", columns = 3) %>%
  add_css_conditional_column(colour_rank_theme = "White-Green", columns = 4) %>%
  add_css_conditional_column(colour_rank_theme = "White-Blue", columns = 5) %>%
  add_css_conditional_column(colour_rank_theme = "White-Red", columns = 6) 
```

## More on tableHTML

It would be impossible to fit all the package's uses on just one article, so I will keep it short here. There are a number of resources / vignettes to have a look at if you would like to find out more about its uses:

* [Tutorial on tableHTML and its use with shiny](https://cran.r-project.org/web/packages/tableHTML/vignettes/tableHTML.html)
* [Conditional formatting](https://cran.r-project.org/web/packages/tableHTML/vignettes/conditional_column.html)
* [Gallery of Examples](https://cran.r-project.org/web/packages/tableHTML/vignettes/examples.html)
* [Latest Version on CRAN](https://cran.r-project.org/web/packages/tableHTML/index.html)

Thanks for reading and as always, if you find any bugs please report them on the [issues](https://github.com/LyzandeR/tableHTML/issues) page on github.

\       

      






